---
title: "Customer Success Analytics"
author: "Jennifer Amador-Gonzalez"
date: "2025-01-03"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load and Explore Data

```{r}
#installying and loading necessary library
library(tidyverse)

#leading dataset from files
file_path <- "/Users/jenniferamador/Downloads/telco.csv"
telco_data <-  read_csv(file_path)

#exploring dataset
head(telco_data) #first few rows
glimpse(telco_data) #structure and data types
summary(telco_data) # summary statistics

#Goal
#check for potential missing values
#understand column data types (categorical vs. numerical)
#identify. the targeet variable (churn) and key predictors (ex: tenure, monthly charges)
        
```

## Clean Data

```{r}
#handle missing values
telco_clean <- telco_data %>% mutate(`Total Charges`=as.numeric(`Total Charges`)) #ensure to convert to numeric
telco_clean <- telco_data %>% drop_na() #drop any rows with missing values

#convert categorical variables to numerical
telco_clean <- telco_data %>% mutate(`Churn` = ifelse(`Churn Label`== "Yes", 1, 0)) #target variable

#Goals
#Handle missing values
#convert categorical variables to numerical for potential modeling
```


##Exploratory Data Analysis (EDA)
```{r}
#load necessary library
library(ggplot2)

#churn distribution
ggplot(telco_clean,aes (x = factor(Churn))) + geom_bar() + labs(title = "Churn Distribution", x = "Churn",  y= "Count")
```

Observations:
The majority of customers (non-churners) do not churn (label 0).
Churners (label 1) make up a smaller proportion of the dataset.
Insight:
This indicates class imbalance, which is common in churn datasets and should be addressed in modeling (e.g., using balanced sampling or weighted metrics).



```{r}
#Correlation heatmap
library(corrplot)
cor_data <- telco_clean %>% select(where(is.numeric))
cor_matrix <- cor(cor_data)
corrplot(cor_matrix, method = "color", type = "upper", tl.col = "black", tl.srt = 45)

# Goals:
# Analyze the distribution of churn.
# Identify trends in numerical variables like tenure, MonthlyCharges, etc.
# Visualize correlations between variables.
```

Objective: The correlation heatmap shows the relationship between numerical features and the target variable (Churn).
Insights:
Tenure in Months has a strong negative correlation with Churn. Employees with longer tenure are less likely to churn.
Monthly Charge shows a positive correlation with Churn. Customers paying higher monthly charges are more likely to churn.
Satisfaction Score and Churn are negatively correlated, indicating dissatisfied employees or customers are more likely to leave.
Churn Score has a strong positive correlation with Churn, suggesting it’s a reliable predictor for attrition.
Actionable Steps:

Focus on improving retention strategies for newer employees/customers.
Evaluate pricing plans for high-paying customers to understand if they perceive value for money.
Improve customer satisfaction by identifying pain points in services.



##Build a Predictive Model

```{r}
# Split data into train and test sets
library(caret)
set.seed(42)
train_index <- createDataPartition(telco_clean$Churn, p = 0.7, list = FALSE)
train <- telco_clean[train_index, ]
test <- telco_clean[-train_index, ]

# Train a logistic regression model
model <- glm(Churn ~ `Tenure in Months` + `Monthly Charge` + `Total Charges`, data = train, family = "binomial")

# Evaluate the model
summary(model)
pred <- predict(model, test, type = "response")
conf_matrix <- table(Predicted = ifelse(pred > 0.5, 1, 0), Actual = test$Churn)
conf_matrix
# Goals:
# Train a simple logistic regression model to predict churn.
# Identify the most important predictors of churn.
```

Observations:
Significant predictors (p < 0.05):
Tenure in Months (negative coefficient): Longer tenure reduces churn likelihood.
Monthly Charge (positive coefficient): Higher charges increase churn likelihood.
Non-significant predictor:
Total Charges: Likely collinear with Tenure in Months.
Insight:
Focus on improving tenure and addressing high monthly charges to reduce churn.


## Data Vizualizations


#Bar Chart: Show churn rates by contract type

```{r}
#Bar Chart: Show churn rates by contract type
library(ggplot2)

# Bar chart of churn rates by contract type
ggplot(telco_clean, aes(x = Contract, fill = factor(Churn))) +
  geom_bar(position = "dodge") +
  labs(title = "Churn Rates by Contract Type", x = "Contract Type", y = "Count") +
  scale_fill_manual(values = c("#00BFC4", "#F8766D"), name = "Churn", labels = c("No", "Yes"))
```

Observations:
Customers with month-to-month contracts churn the most.
Churn rates are much lower for one-year and two-year contracts.
Insight:
Long-term contracts reduce churn, possibly due to commitment and incentives.


#Box Plot: Compare tenure between churned and non-churned customers.

```{r}
# Box plot of tenure by churn
ggplot(telco_clean, aes(x = factor(Churn), y = `Tenure in Months`, fill = factor(Churn))) +
  geom_boxplot() +
  labs(title = "Tenure by Churn", x = "Churn", y = "Tenure (Months)") +
  scale_fill_manual(values = c("#00BFC4", "#F8766D"), name = "Churn", labels = c("No", "Yes"))
```

Observations:
Non-churners (Churn = No) have significantly longer tenure, as indicated by the wider interquartile range and higher median.
Churners (Churn = Yes) have shorter tenure overall.
Insight:
Shorter tenure is a strong indicator of churn, possibly due to dissatisfaction or lack of engagement early in the customer lifecycle.


#Heatmap: Correlation matrix for numerical variables.

```{r}
library(corrplot)

# Calculate correlation matrix
num_data <- telco_clean %>% select(`Tenure in Months`, `Monthly Charge`, `Total Charges`, Churn)
cor_matrix <- cor(num_data)

# Create a heatmap
corrplot(cor_matrix, method = "color", type = "upper", tl.col = "black", tl.srt = 45,
         title = "Correlation Heatmap")
```

Observations:
Tenure in Months is negatively correlated with churn, confirming the insights from the box plot.
Monthly Charge has a slight positive correlation with churn.
Total Charges has a weak correlation with churn, likely because it reflects tenure.
Insight:
Tenure is a key predictor, while Monthly Charge may influence churn when combined with other factors.


# Scatter plot of Monthly Charges vs. Total Charges

```{r}
#ScatterPlot
ggplot(telco_clean, aes(x = `Monthly Charge`, y = `Total Charges`, color = factor(Churn))) +
  geom_point(alpha = 0.6) +
  labs(title = "Monthly Charges vs. Total Charges", x = "Monthly Charges", y = "Total Charges") +
  scale_color_manual(values = c("#00BFC4", "#F8766D"), name = "Churn", labels = c("No", "Yes"))
```

Observations:
A linear trend is visible between Monthly Charge and Total Charges for non-churners and churners.
Churners are distributed across all ranges of monthly charges but seem more concentrated in lower total charges.
Insight:
Customers with lower cumulative charges (shorter tenure) tend to churn more.

# Histogram of Monthly Charges

```{r}
# Histogram
ggplot(telco_clean, aes(x = `Monthly Charge`, fill = factor(Churn))) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  labs(title = "Distribution of Monthly Charges by Churn", x = "Monthly Charges", y = "Count") +
  scale_fill_manual(values = c("#00BFC4", "#F8766D"), name = "Churn", labels = c("No", "Yes"))
```

Objective: The histogram compares the distribution of monthly charges for churned (Yes) vs. non-churned (No) customers.
Insights:
Customers with lower monthly charges (around $20–$40) tend to churn less frequently.
A higher proportion of churned customers are observed in the $70–$100 range.
The distribution suggests that higher charges may be a significant factor in customer attrition.
Actionable Steps:
Offer customized retention plans for high-paying customers.
Consider offering discounts or value-added services to customers in the higher charge brackets.
Investigate if the features/services offered at higher price points meet customer expectations.


